---
title: "01_prcomp_PCA"
author: "Sara Correa Garcia"
date: "02/06/2022"
output: html_document
---
## PCA
Principal component analysis is used to extract the important information from a multivariate data table and to express this information as a set of few new variables called principal components. These new variables correspond to a linear combination of the originals. The number of principal components is less than or equal to the number of original variables.

The information in a given data set corresponds to the total variation it contains. The goal of PCA is to identify directions (or principal components) along which the variation in the data is maximal.
In other words, PCA reduces the dimensionality of a multivariate data to two or three principal components, that can be visualized graphically, with minimal loss of information.

```{r Load libraries}

list.of.packages <- c("ade4","ggfortify", "tidyverse", "factoextra","vegan")  # For data doubs, For function autoplot(ggfortify), For ggplot2 like graphs (factoextra)
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if (length(new.packages) > 0) {
  install.packages(new.packages, dependencies = TRUE) 
  print(paste0("The following package was installed:", new.packages)) 
} else if (length(new.packages) == 0) {
  print("All packages were already installed previously")
}
# Load all packages at once
lapply(list.of.packages, require, character.only = TRUE, quietly = TRUE)

```

### Load and scale data

In principal component analysis, variables are often scaled (i.e. standardized). The goal is to make the variables comparable. Generally variables are scaled to have i) standard deviation one and ii) mean zero.
```{r}
data(doubs)
spe <- doubs$fish
str(spe) 
dim(spe) #30 27
spe = spe[-8,] # We eliminate the site 8, that doesn't contain any species
dim(spe) #29 27
spe.hel = decostand(spe, method = "hellinger") # gives low weights to variables with low counts and many zeros. After a transformation, data may or may not subsequently be standardized. 
```


```{r Split data for PCA and prediction}
set.seed(101) # Set Seed so that same sample can be reproduced in future also
# Now Selecting 75% of data as sample from total 'n' rows of the data  
sample <- sample.int(n = nrow(spe.hel), size = floor(.75*nrow(spe.hel)), replace = F)
train <- spe.hel[sample, ]
test  <- spe.hel[-sample, ]
```


#### Calculating the PCA
the eigenvalues measure the amount of variation retained by each principal component. Eigenvalues are large for the first PCs and small for the subsequent PCs. That is, the first PCs corresponds to the directions with the maximum amount of variation in the data set.

We examine the eigenvalues to determine the number of principal components to be considered. The eigenvalues and the proportion of variances (i.e., information) retained by the principal components (PCs) can be extracted using the function get_eigenvalue() [factoextra package].
```{r PCA}
pca_res <- prcomp(train, center = TRUE, scale. = TRUE) # standardized
fviz_eig(pca_res) #Visualize eigenvalues (scree plot). Show the percentage of variances explained by each principal component.
get_eigenvalue(pca_res)
biplot(pca_res)
summary(pca_res)
names(pca_res)
pca_res$rotation

```
sdev: contains the standard devition of princiapl compononents. We can use sdew to compute the variance explained by each PC. 
rotation: species coordinates
center: contains the mean of our data before scaling
scale : the square of scale contains the variance of each column of our data before scaling
x: stores the principal components 
### Basic graphic

```{r}
var_explained <- pca_res$sdev^2/sum(pca_res$sdev^2)
var_explained[1:5]
pca_res$x %>% 
  as.data.frame %>%
  ggplot(aes(x=PC1,y=PC2)) + geom_point(size=4) +
  theme_bw(base_size=32) + 
  labs(x=paste0("PC1: ",round(var_explained[1]*100,1),"%"),
       y=paste0("PC2: ",round(var_explained[2]*100,1),"%")) +
  theme(legend.position="top")
```


### Graph of sites - Samples with a similar profile are grouped together


```{r}
fviz_pca_ind(pca_res,
             col.ind = "cos2", # Color by the quality of representation
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE     # Avoid text overlapping
             )
```


### Graph of variables
```{r}
fviz_pca_var(pca_res,
             col.var = "contrib", # Color by contributions to the PC
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE     # Avoid text overlapping
             )
```


### Biplot of sites and species/variables
```{r}
fviz_pca_biplot(pca_res, repel = TRUE,
                col.var = "#2E9FDF", # Variables color
                col.ind = "#696969"  # Individuals color
                )
```



### Accessing PCA results:

```{r}
library(factoextra)
# Eigenvalues
eig.val <- get_eigenvalue(pca_res)
eig.val
  
# Results for Variables/Species
res.var <- get_pca_var(pca_res)
res.var$coord          # Coordinates
res.var$contrib        # Contributions to the PCs
res.var$cos2           # Quality of representation 
# Results for Sites
res.ind <- get_pca_ind(pca_res)
res.ind$coord          # Coordinates
res.ind$contrib        # Contributions to the PCs
res.ind$cos2           # Quality of representation 
```


# Predict using PCA

how to predict the coordinates of new sites using only the information provided by the previously performed PCA.

```{r}
site.sup.coord <- predict(pca_res, newdata = test)
site.sup.coord 
```

plot sites including the test subsample/supplementary sites
```{r}
# Plot of active individuals
p <- fviz_pca_ind(pca_res, repel = TRUE)
# Add supplementary individuals
fviz_add(p, site.sup.coord, color = "blue")
```

